---
title: "Untitled"
author: "Tim Riffe"
date: "10/1/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prelims

Given a set of transition probabilities for a state space like this

```{r statespace, echo = FALSE,warning=FALSE,message=FALSE}
tools::texi2dvi(here::here('Figures/statespace.tex'),pdf=TRUE,clean=TRUE)
a<-file.rename(from=here::here("statespace.pdf"),to=here::here("Figures/statespace.pdf"))
#system(paste("cd",getwd(), "&& inkscape -l Figures/statespace.svg Figures/statespace.pdf",sep=" "))
knitr::include_graphics(here::here("Figures/statespace.pdf"))
```

one can calculate the expected time spent in good or poor health given a set if initial conditions. By initial conditions I mean a starting composition at the lowest age, in this case what fraction of people is in good and poor health at the start. 

All we need to get going are the transition probabilities (thanks DCS!) and a once-off calculation of the fundamental matrix *or* an analagous multidecrement (tidy) calculation of the same. Both yield expected occupancy times in each age-state combination conditional on starting in a particular age and state.
```{r, results = "hide", message =FALSE, warning=FALSE}
# helpful for data operations
library(tidyverse)
library(here)
library(reshape2)
# some custom functions
source(here::here("R","Functions.R"))

# Read in the data (Daniel Schneider gave these to me, haha)
TR <- readRDS(here::here("Data","TR_v06_collapsed.rds"))

# define a subset
TRsub <- TR %>% filter(sex == "f",
                       edu == "all_edu",
                       time == 1996)

# starting proportions in each state,
# based on observed prevalence around age 50
init <- TRsub[1,c("s1_prop","s2_prop")] %>% unlist()
names(init) <- c("H","U")

# Make the submatrices of U, the transient matrix
U <- sub2U(TRsub)
# N is the fundamental matrix
N <- U %>% 
  U2N()
```

From this we can calculate the overall expectantcy given that one starts out healthy (`r N[,c("H::48","U::48")] %*% diag(c(1,0)) %>% sum() %>% round(2)`) or unhealthy (`r N[,c("H::48","U::48")] %*% diag(c(0,1)) %>% sum() %>% round(2)`) at age 50. This is fine to report, and sometimes there are motives to make statements about expected outcomes given a particular starting condition. However, if we wish to make a statement about the population expectancy, then these two quantitites must somehow be blended. One could

 - Measure prevalence at the youngest age at a given time: `r N[,c("H::48","U::48")] %*% diag(init) %>% sum() %>% round(2)`
 - Assume constant transition probabilities in all ages before the youngest age: `r N[,c("H::48","U::48")] %*% diag(init_constant(TRsub)) %>% sum() %>% round(2)`
 - Assume a stationary process: `r N[,c("H::48","U::48")] %*% diag(c(.9158685, 1-.9158685)) %>% sum() %>% round(2)` (work in progress)
 - Extrapolate transition probabilties to younger ages, assume one of the above, then project forward to the starting age. (not done here)

It makes very little difference in this case, as concerns overall expectancy. The leverage of making a good choice of initial conditions depends on: mortality differences between states, and turnover between states. High mortality differentials make initial conditions matter more, but fast mixture (high onset and recovery) makes intial conditions matter very little.

To get a sense, here are those same expectancies if transient turnover (onset and recovery) is reduced by half, but mortality stays the same:

```{r include = FALSE}
# where mat is just transitions originating in a state
# hard coded to this particular data.... careful!

# N is the fundamental matrix
Nlow <- TRsub %>% 
  rescale_turnover_U(.5) %>% 
  U2N()
```

So this is why I want to explore different options of generating reasonable initial conditions. We will compare the *constant assumption* with a derivation of stationary initial conditions.

The idea of stationary initial conditions comes from Brouard 

```{r, echo = FALSE}
x <- seq(48,110,by=2)

# occupancy times (survival curve) depending on health status at start
hsx <- N[1:32,"H::48"]+N[33:64,"H::48"]
usx <- N[1:32,"U::48"]+N[33:64,"U::48"]
# prevalence healthy by age depending on health status at start
hin <- N[1:32,"H::48"] / hsx
uin <- N[1:32,"U::48"] / usx
```

```{r, echo = FALSE}
plot(x, hin, type= 'l')
lines(x, uin)
polygon(c(x,rev(x)),c(hin,rev(uin)),col = "#9E87BB")
```

```{r}
plot(x,hsx/2,type = 'l')
lines(x,usx/2)
sum(hsx) - sum(usx)
```

# set up simulation

Let's at least do a test where we range init from 0 to 1 in steps of .01, and where we range turnover from .1 to 10 in evenly spaced log10 steps.

The way turnover is scaled merits some explanation. 
```{r, echo = FALSE}



plot_turnover(TRsub,1)
plot_turnover(TRsub,.5)
```

```{r}
library(colorspace)
library(ggplot2)

init_steps     <- seq(0,1,by=.01)
turnover_steps <- exp(seq(log(.1),log(10),length=101))

e50mat <- matrix(NA,
                nrow = length(init_steps),
                ncol = length(turnover_steps),
                dimnames = list(init_steps = init_steps,
                                turnover_steps = turnover_steps))

for (j in 1:length(turnover_steps)){
  Ni <- rescale_turnover_U(TRsub, turnover_steps[j]) %>% 
   U2N()
  
  for (i in 1:length(init_steps)){
    e50ij <- Ni[,c("H::48","U::48")] %*% diag(c(init_steps[i], 1 - init_steps[i])) %>% sum()
  e50mat[i,j] <- e50ij
  }
}
# image(init_steps,
#       turnover_steps,
#       e0mat,
#       xlab = "Fraction unhealthy at age 50",
#       ylab = "log10(turnover_steps)",
#       log = 'y')
# contour(init_steps,
#         turnover_steps,
#         e0mat,
#         add = TRUE,
#         log = 'y',
#         cex=3)
N <- sub2U(TRsub) %>% U2N()
e50const <- 
  N[,c("H::48","U::48")] %*% diag(init_constant(TRsub)) %>% 
  sum()
e50diff <- e50mat - e50const


e50diff %>% 
  reshape2::melt() %>% 
  rename(e50diff = value) %>% 
  ggplot() + 
  aes(x = turnover_steps, y = init_steps, z = e50diff) + 
  scale_x_log10()+
  geom_tile(aes(fill=e50diff)) + 
  stat_contour(aes(fill=..level..), geom="polygon") + 
  #geom_contour(color="white", alpha=0.5) +
  scale_fill_continuous_diverging("Blue-Red") + 
  theme_bw() + 
  labs(title = "LE diff by turnover scale and initial fraction healthy")+ 
  xlab("turnover scale factor") + 
  ylab("initial fraction healthy") +
  annotate(geom = "point",x=1,y=init_constant(TRsub)[1])
```

# back-project transitions, then forward project starting prev.

Log-linear back-extrap of mortality.
```{r}
logit <- function(x){log(x / (1 - x))}
expit <-  function(x){exp(x) / (1 + exp(x))}

init_back_proj(TRsub, 
               start_age = 40,
               max_age_fit = 65,
               interval = 2)
```



